\documentclass[a4paper,12pt]{article}

\usepackage{listings}
\usepackage[utf8]{inputenc}

\usepackage[margin=1.00in]{geometry}
\usepackage{tikz-cd}
\usepackage{cite}
\usepackage{titling}
\usepackage{indentfirst}
\usepackage{amsmath}
\usepackage[textwidth=3.7cm]{todonotes}
\usepackage{faktor}

\setlength{\marginparwidth}{3.7cm}
\usepackage{etoolbox}
\lstset{breaklines}

\usepackage[compact]{titlesec}         % you need this package
\titlespacing{\section}{0pt}{0pt}{0pt} % this reduces space between (sub)sections to 0pt, for example
\titlespacing{\subsection}{0pt}{0pt}{0pt} % this reduces space between (sub)sections to 0pt, for example
\titlespacing{\subsubsection}{0pt}{0pt}{0pt} % this reduces space between (sub)sections to 0pt, for example
\AtBeginDocument{%                     % this will reduce spaces between parts (above and below) of texts within a (sub)section to 0pt, for example - like between an 'eqnarray' and text
    \setlength\abovedisplayskip{-12pt}
    \setlength\belowdisplayskip{3pt}
}

\setlength{\droptitle}{-5em} % This is your set scream
\setlength{\parindent}{1em}
\setlength{\parskip}{0.8em}
\setlength{\parindent}{0in}
\date{}

\title{Temp\vspace{-3.5em}}

\begin{document}
\maketitle
A common notion in Category theory is that of a universal construction.
Universal constructions are common patterns that occur throughout mathematics
that aim to capture the essence of these patterns at the categorical level.
Before attempting to abstract these ideas it is useful to consider some
examples. Only the constructions most relevant to the heart of the thesis will
be discussed here.  After looking at the individual constructions we will look
at three ways of abstracting them and look at the links between them.

\section{Products}
Products exemplify a common construction in categories of combining the
structure of two objects (in some canonical way) within the category to produce
an object of the same category. In more concrete terms the product of two
objects $A$ and $B$ in the category $\textbf{C}$ is an triple $(A \times B,
\pi_{1}, \pi_{2})$ where for all other objects $C$ in $\textbf{C}$ with
projections $f: C \rightarrow A$ and $g: C \rightarrow B$ the unique arrow
$\langle f, g\rangle : C \rightarrow A \times B$ can be formed such that the
following diagram commutes (meaning that every path through the diagram is
equal).

\[\begin{tikzcd}[sep=huge]
 & C \arrow[ld, "f"'] \arrow[rd, "g"] \arrow[d, "{\langle f, g\rangle}" description, dashed] &  \\
A & A \times B \arrow[l, "\pi_2"] \arrow[r, "\pi_1"'] & B
\end{tikzcd}\]

where the dashed arrow indicates uniqueness. This can be extended to
\textit{n}-ary products in the obvious way.

As is common with universal constructions, products (via the unique arrow
$\langle f, g \rangle$) are unique up to unique isomorphism.

Examples of products within familiar categories include the cartesian product
$\times$ in \textbf{Set}, defined as the set of all tuples of elements from two
separate sets.

If products can be formed for every finite set of objects in a category it is
said to be cartesian.

\section{Terminal Objects}
Terminal objects are constructions that capture the
minimal structure required to be an object within a category. They often
correspond to the trivial examples of certain constructions.

A terminal object of a category \textbf{C} is an object, T, such that, for all
other objects, $A$ in the category, there exists a unique arrow $!_{A}: A
\rightarrow T$.

As a commutative diagram:

\[\begin{tikzcd}[sep=huge]
A \arrow[d, "!_A", dashed] \\
T
\end{tikzcd}\]

As with products, terminal objects in categories are unique up to unique
isomorphism.

Examples of terminal objects in common categories include any singleton in
\textbf{Set} and the one element group in \textbf{Group}


\section{Universal Properties}
As is the norm in category theory, if a common pattern is found the aim is to
abstract it to find its underlying mechanism. Universal properties can
be abstracted as initial or terminal morphisms.

\subsection{Initial Morphisms}
An initial morphism is an initial object in the category $X \downarrow U$ where
$U : C \rightarrow D$ is a functor and $X$ is an object in $C$. More precisely,
an initial morphism is a pair $(A, \varphi)$ such that the following diagram
commutes.

A terminal morphism is a terminal object in the category $U \downarrow X$

\[\begin{tikzcd}
    A \arrow[rd] \arrow[r, "\phi"] & B \\
                                   & C
\end{tikzcd}\]

\section{Adjunctions}
\section{Yoneda Lemma and Embedding}
The Yoneda lemma is a core result in Category Theory. In a liberal sense it can
be thought to justify that by examining the ways in which an object can be
manipulated we can totally determine the object. This is a consequence of the
Yoneda Embedding, a functor from a (locally small) category $C$ to the category of
\textit{presheaves} (functors of type : $C \rightarrow \textbf{Set}$) on $C$.
The Yoneda lemma guarantees that this functor is fully faithful and injective on
objects, allowing the category of presheaves to be worked with in place of the
category itself. The complexity of this theorem in comparison to other
categorical constructs motivates a full treatment below.

Concretely the Yoneda lemma is the following statement:

\begin{equation}
    Nat(h^{A}, F) \cong FA
    \label{Yoneda}
\end{equation}

and that this isomorphism is natural in both $A$ and $F$.

The left hand side of \ref{Yoneda} can be understood as follows. $\phi : Nat(h^{A},
F)$ are natural transformations between the covariant hom-functor and some other
functor $F$, both presheaves on some underlying category $C$. Components of
$\phi$ at $X$ take morphisms from $A$ to $X$ to elements of the set $FX$.

Naturality will be dealt with after first establishing the isomorphism in
\ref{Yoneda}.

To establish the isomorphism, two functions must be defined. $Y: Nat(h^{A}, F)
\rightarrow FA$ and $Y^{op}: FA \rightarrow Nat(h^{A}, F)$ that are mutual
inverses.

The two functions are (using $\lambda$ abstraction for convenience):

% Y(\phi) = \phi_{A}(id_{A})
% Y^{op}(u) = \lambda f . F f u
\begin{align*}
    Y(\phi) = \phi_{A}(id_{A}) \\
    Y^{op}(u) = \lambda f . F f u
\end{align*}

$Y \circ Y^{op}$ can be shown to be $id$ as follows:

\begin{align*}
    (Y \circ Y^{op}) \ u &= Y (\lambda f \,.\, F f \ u) \\
    &= (\lambda f \,.\, F f \: u ) \ id_{A} \\
    &= (F \ id_{A}) \ u \\
    &=  id_{FA} u \\
    &= u
\end{align*}

To show $Y^{op} \circ Y \ = \ id$ proceed with:

\begin{align*}
    (Y^{op} \circ Y) (\phi) &= Y^{op} (\phi_{A}(id_{A})) \\
    &= \lambda f \, . \, F f \ (\phi_{A}( id_{A}))) \\
    &= \lambda f \, . \, \phi_{Y}(f) \\
    &= \phi
\end{align*}


Where the second to third line come from the fact that $\phi$ is a natural
transformation such that the following diagram commutes.

\[\begin{tikzcd}[sep=huge]
    {Hom(A, X)} \arrow[d, "{Hom(A,f)}"'] \arrow[rr, "\phi_{X}"] &  & FX
    \arrow[d, "Ff"] \\
    {Hom(A,Y)} \arrow[rr, "\phi_{Y}"] &  & FY
\end{tikzcd}\]

Now it must be shown that $Y^{op}(u)$ is a natural transformation i.e for some
$g: A \rightarrow Z$: 

\begin{align*}
    \lambda f \, . \, F f u  \circ hom(A, g) = F g \circ \lambda f \, . \, F f u
\end{align*}

Taking the left hand side on some $h: A \rightarrow X$:

\begin{align*}
    (\lambda f \, . \, F f u  \circ hom(A, g)) \ h &= (\lambda f \, . \, F f u) \ (
    hom(A, g) \ h) \\
    &= (\lambda f \, . \, F f u) \ (g \circ h) \\
    &=  F \ (g \circ h) \ u
\end{align*}

And now the right hand side:

\begin{align*}
    (F g \circ \lambda f \, . \, F f u) \ h &= F \ g \ ( (\lambda f \, . \, F f
        u) \
    h) \\
    &= F \ g \ (F \, h \, u) \\
    &= F \ (g \circ h) \ u
\end{align*}

Thus establishing $Y^{op}(u)$ as a natural transformation.

Finally naturality in $A$ and $F$ in \ref{Yoneda} must be established.

To establish naturality in $A$ it must be shown that, for all $f: C \rightarrow D$, the
following diagram commutes:

\[\begin{tikzcd}[sep=huge]
    {hom(h^C, F)} \arrow[d, "{hom(h^f, F)}"'] \arrow[rr, "Y"] &  & {F \,C}
    \arrow[d, "{F \, f}"] \\
    {hom(h^D, F)} \arrow[rr, "Y"] &  & {F \, D}
\end{tikzcd}\]

i.e.

\begin{align*}
    F \, f \circ Y = Y \circ hom(h^{f}, F)
\end{align*}

Beginning with the left hand side

\begin{align*}
    F \, f \ ( Y(\phi)) &= F \, f \ (\phi \ (id_{C})) \\
    &= (F \, f \circ \phi) \ (id_{C}) \\
    &= (\phi \circ h^{A}(f)) \ (id_{C}) \\
    &= \phi \ ((h^{A}(f) \ (id_{C})) \\
    &= \phi \ (f \circ id_{C}) \\
    &= \phi \, f
\end{align*}

And now the right

\begin{align*}
    (Y \circ hom(h^{f}, F)) \ (\phi) &= Y \ (hom(h^{f}, F) \, (\phi)) \\
    &= Y \, (\phi \circ h^{f}) \\
    &= (\phi \circ h^{f}) \ (id_{C}) \\
    &= \phi \ (h^{f} \, (id_{C})) \\ 
    &=  \phi \, f
\end{align*}

A contravariant version of Yoneda can be proven which states the following:

\begin{equation}
    Nat(h_{A}, F) \cong FA
    \label{coyoneda}
\end{equation}

$F$ can be fixed to certain functors to examine what Yoneda lemma means in
certain contexts.

Using the contravariant Yoneda and setting $F$ to be $h_{B}$ the equation

\begin{equation}
    Nat(h_{A}, h_{B}) \cong hom(A,B)
    \label{yonembedding}
\end{equation}


An obvious functor from any category \textbf{C} to the categor of presheafs from
\textbf{C} assigns the functor $h^{A}$ for all objects $A:\textbf{C}$ and
assigns the natural transformation $h^{f}$ to all morphisms $f$ in \textbf{C}.

It can be shown, using the Yoneda lemma, that this functor is fully faithful on
arrows and injective on objects.

\section{Category of Boolean Algebras}
A boolean algebra is a partial order (reflexive and transitive), $(S, \leq)$ such that that
there exist three operations $\land, \lor: S\times S \rightarrow S$  and $\neg:
S \rightarrow S$ known as meet, join and complementation respectively such that
the folliwng equations hold.

\begin{align*}
    &\exists \ \textbf{0}, \textbf{1} \ s.t. \ \forall \ c \in S. \ \textbf{0}
    \leq c \, , \ c \leq \ \textbf{1} \\
    &\forall \, a, b, c \in S. \ a \leq c \, , \ b \leq c \ \Leftrightarrow \ a \,
    \lor \, b \leq c \\
    &\forall \, a, b, c  \in S. \ c \leq a \, , \ c \leq b \ \Leftrightarrow \ c
    \leq a \, \land \, b \\
    &\forall \, a, b \in S. \ a \leq \neg \, b \ \Leftrightarrow \ a \, \land \,
    b = 0 \\
    &\forall \, a \in S. \   a = \neg \, \neg \, a
\end{align*}

Some familiar objects in the category of boolean algebras include the powerset
of a set $\mathcal{P}(A)$ with the subset relation $\subseteq$. This object has
its meet as $\cap$, its join as $\cup$, its complementation as set
complementation $A^c$, the \textbf{0} as $\emptyset$, the \textbf{1} as the set
itself, $A$.

The initial object in this category is the familiar two-element boolean algebra
\textbf{2} which contains $ \{\textbf{0}, \textbf{1}\}$ under the normal logical
interpretation of the operators $\land$, $\lor$ and $\neg$ under the obvious
ordering.

\todo{products as meet}

Morphisms into the initial object are precisely the ultrafilters on a boolean
algebra.


\section{$\omega$-CPOs}

\section{Exponential Objects}
Exponential objects are universal constructions that capture the notion of
function spaces or higher order objects.

The object $B^{A}$ indicates the mappings from the object $A$ to $B$. This is
paired with the morphism $eval: B^{A} \times A \rightarrow B$ such that for any
object $Z$ and morphism $f : Z\times A \rightarrow B$ there exists a unique
morphism $\tilde{f}: Z \rightarrow B^{A}$ such that the following diagram
commutes:

\todo{exponential diagram}

$\tilde{f}$, or transposition, can be thought of as currying in the functional
programming sense, taking a function in multiple arguments to a sequence of
functions in one argument.

If exponents can be formed for every pair of objects in a category, \textbf{C},
exponentiation under a given object, $A$, forms a functor $\_^{A}: \textbf{C}
\rightarrow \textbf{C}$.

$\_^{A}$ takes an object, $B$, to the exponential object, $B^{A}$ and arrows
$\alpha: C \rightarrow D$ to the transposition of $\beta \circ eval$.

For any boolean algebra regarded as a poset category itself we have the
exponential $B^{A}$ defined as the object $\neg \, A \ \lor B$ otherwise
understood as the implication $A \implies B$
\section{Heyting Algebras}
\section{Equalisers}
Equalisers are universal constructions that correspond to notions such as the
kernel of homomorphisms and the objects that are mapped to the same place by
multiple arrows.

Given arrows $f, g : A \rightarrow B $ the equaliser of the two arrows an object
$E$ and the morphism $e: E \rightarrow A$ such that the following diagram
commutes:

\[\begin{tikzcd}[sep=huge]
    E \arrow[r, "e"] & A \arrow[r, "f", bend left] \arrow[r, "g"', bend right] &
    B
\end{tikzcd}\]

\[\begin{tikzcd}[sep=huge]
    B^A \times A \arrow[r, "eval"]                         & B \arrow[r, "g"]
    & B \\
    A \times A \arrow[u, "\langle \phi \times id \rangle"] &
    &   \\
    A \arrow[u, "\delta", no head]                         & 1 \arrow[l, "u"]
    \arrow[ruu, "\phi \ \circ \ u \ \circ \ u"'] &  
\end{tikzcd}\]


Equalizers can be used to specify subsets via the following construction:

For each subset, $U$ of a set $A$ we can form the the canonical mapping
$\chi_{U}: X \rightarrow \textbf{2}$ which takes all the elements of $U$ in
$X$ to \textbf{1} and all elements of $X$ not in $U$ to \textbf{0}.

The inclusion mapping $i: U \rightarrow A$ equalises the above defined
$\chi_{U}$ and the the constant map from $A$ to the $\textbf{1} \in \textbf{2}$.

An inverse, $V$ can be produced for the mapping $\chi$ showing that these
functions are in a one-to-one mapping with the subsets of $X$ establishing an
isomorphism between the functions from $A \rightarrow \textbf{2}$ and the
powerset of $A$.

The, $V$, for a function of type $A \rightarrow \textbf{2}$ returns the set of
objects in $A$ that map to the value \textbf{1} i.e. precisely the elements of
$U$. This function is an equaliser similar to $i$ above.

The correspondence between equalisers and subsets can be made more clear by
first acknowledging the fact that equalisers are monic. This can be seen by
establishing that if an object, $Z$ has an arrow, $g$ to the equaliser $E$, this
must be unique, as the composition of this arrow with $e$ equalises the two
arrows that $E$ equalises. This induces the property that there exists a unique
arrow from $Z$ to $E$ so any two arrows to $E$ must in fact be the same arrow.

\section{Subobjects}

Given that subsets coincide with concept of equalisers and that equalisers are
all mono, within categories without explicit notions of subset relations,
monomorphisms can be regarded as an abstrac subset notion. When viewed in this
light, monomorphisms are called subobjects. A monomorphism $m: A \rightarrow X$
is called a subobject of $X$.

Subobjects ($A \rightarrow X)$ can themselves be compared to each other by considering the
\todo{work out quotient notation} slice category of arrows to $X$.

Arrows between subobject domains are morphisms in this slice category and can be
viewed as subset relations themselves.
\end{document}
