% !TEX root = ./dissertation.tex
\vspace{1cm}
\noindent
The motivating force behind this thesis was an investigation into mathematical
logic with a particular focus in theorem proving and category theory.

~\\
Within the thesis an account of category theory has been presented within the
theorem prover Agda. Lawvere's fixed point theorem, a significant theorem
within the study of mathematical logic and  category theory has been
formalised. Proofs in the theorem prover Agda are detailed step by step and the
subtleties of the language explained.

~\\
Two applications have been provided: The first, a type-theoretic analogue of
Russell's paradox within a constructed category of small types. The second, a
more in depth attempt, was a novel application of Lawvere's fixed point theorem
to the untyped $\lambda$-calculus. This was achieved by examining the
relationship between models of the $\lambda$-calculus and the theory of
cartesian closed categories and resulted in the observation that the Lawvere's
fixed point theorem could be used to derive the first fixed point theorem in
the untyped $\lambda$-calculus.

~\\
An analysis of the limitations within theorem proving have been provided and
explained alongside a presentation of modern advancements within type theory
that rectify this.

~\\
The importance of theorem proving as a mathematical practice has been explained
with evidence to suggest its worth. The interwoven historical context behind
both theorem proving and Lawvere's fixed point theorem have been examined and
the theory that underpins this has been presented.

~\\
This dissertation consists of the explanations herein and an online repository
of fully mechanised proofs.
