\documentclass[ % the name of the author
                author={Alessio Zakaria},
                % the name of the supervisor
                supervisor={Dr. Nicolas Wu},
                % the degree programme
                degree={MEng},
                % the dissertation title (properly capitalised)
                title={Automated Theorem Proving in Category Theory and the
                $\lambda$-calculus},
                % the dissertation subtitle (which can be blank)
                subtitle={},
                % the dissertation type
                type={Research},
                % the year of submission
                year={2019} ]{dissertation}


\begin{document}


% =============================================================================

% This macro creates the standard UoB title page by using information drawn
% from the document class (meaning it is vital you select the correct degree
% title and so on).

\maketitle

% After the title page (which is a special case in that it is not numbered)
% comes the front matter or preliminaries; this macro signals the start of
% such content, meaning the pages are numbered with Roman numerals.

\frontmatter

% This macro creates the standard UoB declaration; on the printed hard-copy,
% this must be physically signed by the author in the space indicated.
\makedecl

% LaTeX automatically generates a table of contents, plus associated lists
% of figures, tables and algorithms.  The former is a compulsory part of the
% dissertation, but if you do not require the latter they can be suppressed
% by simply commenting out the associated macro.

\tableofcontents

% The following sections are part of the front matter, but are not generated
% automatically by LaTeX; the use of \chapter* means they are not numbered.

% -----------------------------------------------------------------------------
\mainmatter

% -----------------------------------------------------------------------------
\chapter*{Acknowledgements}
I would like to thank my supervisor, Dr. Nicolas Wu for his continued enthusiasm
and inspiration. I would like to thank Dr. David Bernhard for enabling me to
become a mathematician. I would like to thank Csongor Kiss and Jannis Limperg
for their guidance though all things types and Agda. Thank you to my friends and
family for being everything.
\chapter*{Executive Summary}

The research hypothesis of this thesis is that there is a significant advantage
to employing computer-aided proof assistants in the study of category theory and
mathematical logic. This was done by implementing a theorem within the study of
categorical logic in the theorem prover Agda. By doing this, the limitations and
importance of theorem proving was explored in a practical way.

\begin{itemize}
    \item I learnt the underlying theory of modern theorem provers
    \item I learnt the practical usage of the theorem prover Agda
    \item I formalised Lawvere's fixed point theorem within Agda
    \item I obtained a novel application of the theorem within the untyped
        $\lambda$-calculus
    \item I surveyed modern advancements in type theory that eliminate the
        limitations of modern theories.
\end{itemize}

\chapter*{Supporting Technologies}

This thesis made use of two supporting technologies

\begin{itemize}
    \item version 2.5.4.2 of the theorem prover Adga to formalise theorems
        within Category Theory
    \item the \verb|cats| library by Jannis Limperg \cite{limperg2019cats}
\end{itemize}
\chapter{Contextual Background}
\label{chap:context}



\input{context.tex}

% -----------------------------------------------------------------------------

\chapter{Technical Background}
\label{chap:technical}

\input{technical.tex}

% -----------------------------------------------------------------------------

\chapter{Project Execution}
\label{chap:execution}

\input{execution.tex}

% -----------------------------------------------------------------------------

\chapter{Critical Evaluation}
\label{chap:evaluation}

\input{evaluation.tex}

% -----------------------------------------------------------------------------

\chapter{Conclusion}
\label{chap:conclusion}

\input{conclusion.tex}

% =============================================================================

% Finally, after the main matter, the back matter is specified.  This is
% typically populated with just the bibliography.  LaTeX deals with these
% in one of two ways, namely
%
% - inline, which roughly means the author specifies entries using the
%   \bibitem macro and typesets them manually, or
% - using BiBTeX, which means entries are contained in a separate file
%   (which is essentially a database) then imported; this is the
%   approach used below, with the databased being dissertation.bib.
%
% Either way, the each entry has a key (or identifier) which can be used
% in the main matter to cite it, e.g., \cite{X}, \cite[Chapter 2}{Y}.

\backmatter

\bibliography{dissertation}

% -----------------------------------------------------------------------------

% The dissertation concludes with a set of (optional) appendicies; these are
% the same as chapters in a sense, but once signaled as being appendicies via
% the associated macro, LaTeX manages them appropriatly.

\appendix

\chapter{The Karoubi Map and Lawvere's Fixed Point Theorem}
\label{appx:karoubi}

Below is a failed attempt to see the consequences of lawvere's fixed pont
theorem after applying the Karoubi Envelope to the $\lambda$-algebra.

As stated earlier every $\lambda$-algebra can be turned into a ccc with a
reflexive object via the karoubi envelope which turns any additive category into
a pseudoabelian category. Let $C = (X, \cdot , k, s)$ be a $\lambda$-algebra the
karoubi envelope is as follows where $a \circ b = \lambda x . a (b x)$.

\begin{align*}
    &\textnormal{Objects}: \, \{ x \in X \, | \, x \circ x = x \} \\
    &\textnormal{Arrows}: \, \textnormal{Hom}(a, \, b) =  \{ f \in X \, | \, b
        \circ f
    \cdot a = f\} \\
    &\textnormal{Identity}: \, \textnormal{id}_{a} = a \\
    &\textnormal{Composition}: \, f \circ g
\end{align*}

This comes from considering $C$ as a monoid which is then categorified as a
category with a single object.

Proofs of the validity of all constructions can be found in (Koymans 1982). To
show cccsness.

\begin{align*}
    &\textnormal{Terminal}: \, t = \lambda x y . y \\
\end{align*}
\ldots

The reflexive object is \textbf{I} as $\textbf{I}^\textbf{I} = \textbf{1}$ where
$\textbf{I} = \lambda x.x$ and $\textbf{1} = \lambda x \, y . x \, y$ where
\textbf{1} is itself the arrow both ways between \textbf{1} and \textbf{I} and
$\textbf{1} \circ \textbf{1} = \textnormal{id}_{\textbf{1}}$.

Via the earlier corollary, every endomorphism on \textbf{I} has a fixed point.
Endomorphisms on \textbf{I} are $\lambda$-terms, $f$, such that $\textbf{I}
\circ f \circ \textbf{I} = f$. Expanding this out
\begin{align*}
    \textbf{I} \circ f \circ \textbf{I} &= \lambda x . \, \textbf{I} \, (f \, (\textbf{I}
    \, x)) \\
    &= \lambda x . \, f \, x
\end{align*}

For this to encompass all (any?) $\lambda$-terms our equational theory must
include the $\eta$-rule. Given that this is a construction is for
$\lambda$-algebras which need not have $\eta$ this rains on the hope that
Lawvere's fixed point theorem will correspond to the first recursion theorem for
the untyped $\lambda$-calculus.

A point to \textbf{I}, $p$, in the karoubi'd category corresponds to $\lambda$-terms such
that $\textbf{I} \circ p \circ t = p$

\begin{align*}
    \textbf{I} \circ p \circ t &= p \circ t \\
    &= \lambda x . p (t x) \\
    &= \lambda x . p (\lambda y . y) \\
    &= \lambda x . p \textbf{I} \\
    &= \textbf{K} (p \textbf{I})
\end{align*}

i.e some constant function.

With this in mind, Lawvere's fixed point theorem ends up representing a
reasonably strange theorem in the untyped $\lambda$-calculus i.e. for every
$\lambda$-term, $f$, that satisfies extensionality there exists a constant
$\lambda$-term, $u$, such that $\lambda x . f (u x) = u$. $u =
\textbf{K}(p\textbf{I})$ for some $p$.

\begin{align*}
    \lambda x . f (u x ) &= \lambda x . f(( \textbf{K} (p \textbf{I})) x) \\
    &= \lambda x . f(p \textbf{I}) \\
    &= \textbf{K} (f (p \textbf{I})) \\
    &= \textbf{K} (p \textbf{I})
\end{align*}


% =============================================================================

\end{document}
